\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-17 jasa/paper-jasa.tex
scikitlearn and gpyopt

In this section, we illustrate the proposed method on two representative
examples where the exact likelihood is known and then discuss its application to high energy physics. The code used to produce the results and extended details for these examples is available in Ref.~\citep{carl}.%\footnote{\url{https://github.com/diana-hep/carl}}.

@@

Figure~\ref{fig:5a} shows the exact log-likelihood ratio for this dataset, which has an exact MLE at $(\hat \alpha=1.012, \hat \beta=-0.9221)$.
Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values. Some roughness in the contours is observed, which is primarily due to variance introduced in the calibration procedure. In addition to the statistical fluctuations due to finite calibration samples, there are also fluctuations introduced from changes in the binning of the calibration histograms as $\alpha$ and $\beta$ vary. As discussed in Sec.~\ref{sec:param-calibration}, a parameterized calibration
procedure should ameliorate this issue, but that is left for now as an area for future work.
Nevertheless, optimizing the approximate log-likelihood ratio with
a Bayesian optimization~\citep{brochu2010tutorial} procedure is efficient and effective.
After $50$ likelihood evaluations, the maximum likelihood estimate is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$.
While the objective of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the
internal Gaussian process, shown in Fig.~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Fig.~\ref{fig:5a}.
In each case, the true values $\alpha=1$ and $\beta=-1$ are contained within the $1-\sigma$ likelihood contour.
%For comparison, Figure~\ref{fig:5b} shows the approximate log likelihood ratio evaluated on a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
%roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.

@@

In this work, we have outlined an approach to reformulate generalized likelihood
ratio tests with a high-dimensional data set in terms of univariate densities
of a classifier score. We have shown that a parameterized family of
discriminative classifiers $\hat s(\mathbf{x}; \theta_0, \theta_1)$ trained and
calibrated with a simulator can be used to approximate the likelihood ratio,
even when it is not possible to directly evaluate the likelihood
$p(\mathbf{x}|\theta)$.
% A technique for decomposing this ratio when the generative model is a mixture
% of components was presented with the aim to help focus capacity of the
% classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small
% mixture coefficient.  This approach leverages the power of machine learning in a
% classical statistical setting.
The proposed method offers an alternative to Approximate Bayesian Computation
for parameter inference in the likelihood-free setting that can also be used in
the frequentist formalism without specifying a prior over the parameters. 
In contrast to approaches that learn the posterior conditional on $\mathcal{D}$, our
approach can be applied to any observed data $\mathcal{D}$ once trained.
A strength of this approach is that it separates the quality of the approximation
of the target likelihood from the quality of the calibration. The former is
leverages the continuing advances in supervised learning approaches to classification. 
The calibration procedure for a particular
parameter point is fairly straightforward since it involves estimating a
univariate density using a generative model of the data. The difficulty of the
calibration stage is performing this calibration continuously in $\theta$.
Different strategies to this calibration are anticipated depending on the
dimensionality of $\theta$, the complexity of the resulting likelihood function,
or the practical issues associated to running the simulator.

@@

For $P(Y=0)=P(Y=1)=\frac{1}{2}$, the best regression function $s^*$ simplifies
to
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
\end{equation}
If we further assume that samples for $Y=0$ (resp. $Y=1$) are drawn from some parameterized
distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$ (resp. $p_{\mathbf{X}}(\mathbf{x}|\theta_1)$), then the best regression function can be rewritten
as
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
In particular, this regression function satisfies conditions of
Thm.~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic with
$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.

>>>>

In this section, we illustrate the proposed method on two representative
examples where the exact likelihood is known and then discuss its application to high energy physics. The code used to produce the results and extended details for these examples is available in Ref.~\citep{carl}, which utilizes the classification and calibration routines in scikit-learn~\citep{scikit-learn}. %\footnote{\url{https://github.com/diana-hep/carl}}.

@@

Figure~\ref{fig:5a} shows the exact log-likelihood ratio for this dataset, which has an exact MLE at $(\hat \alpha=1.012, \hat \beta=-0.9221)$.
Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values. Some roughness in the contours is observed, which is primarily due to variance introduced in the calibration procedure. In addition to the statistical fluctuations due to finite calibration samples, there are also fluctuations introduced from changes in the binning of the calibration histograms as $\alpha$ and $\beta$ vary. As discussed in Sec.~\ref{sec:param-calibration}, a parameterized calibration
procedure should ameliorate this issue, but that is left for now as an area for future work.
Nevertheless, optimizing the approximate log-likelihood ratio with
a Bayesian optimization~\citep{brochu2010tutorial,gpy2014} procedure is efficient and effective.
After $50$ likelihood evaluations, the maximum likelihood estimate is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$.
While the objective of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the
internal Gaussian process, shown in Fig.~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Fig.~\ref{fig:5a}.
In each case, the true values $\alpha=1$ and $\beta=-1$ are contained within the $1-\sigma$ likelihood contour.
%For comparison, Figure~\ref{fig:5b} shows the approximate log likelihood ratio evaluated on a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
%roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.

@@

In this work, we have outlined an approach to reformulate generalized likelihood
ratio tests with a high-dimensional data set in terms of univariate densities
of a classifier score. We have shown that a parameterized family of
discriminative classifiers $\hat s(\mathbf{x}; \theta_0, \theta_1)$ trained and
calibrated with a simulator can be used to approximate the likelihood ratio,
even when it is not possible to directly evaluate the likelihood
$p(\mathbf{x}|\theta)$.
% A technique for decomposing this ratio when the generative model is a mixture
% of components was presented with the aim to help focus capacity of the
% classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small
% mixture coefficient.  This approach leverages the power of machine learning in a
% classical statistical setting.
The proposed method offers an alternative to Approximate Bayesian Computation
for parameter inference in the likelihood-free setting that can also be used in
the frequentist formalism without specifying a prior over the parameters. 
In contrast to approaches that learn the posterior conditional on $\mathcal{D}$, our
approach can be applied to any observed data $\mathcal{D}$ once trained.
A strength of this approach is that it separates the quality of the approximation
of the target likelihood from the quality of the calibration. The former
leverages the continuing advances in supervised learning approaches to classification. 
The calibration procedure for a particular
parameter point is fairly straightforward since it involves estimating a
univariate density using a generative model of the data. The difficulty of the
calibration stage is performing this calibration continuously in $\theta$.
Different strategies to this calibration are anticipated depending on the
dimensionality of $\theta$, the complexity of the resulting likelihood function,
or the practical issues associated to running the simulator.

@@

For $P(Y=0)=P(Y=1)=\sfrac{1}{2}$, the best regression function $s^*$ simplifies
to
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
\end{equation}
If we further assume that samples for $Y=0$ (resp. $Y=1$) are drawn from some parameterized
distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$ (resp. $p_{\mathbf{X}}(\mathbf{x}|\theta_1)$), then the best regression function can be rewritten
as
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
In particular, this regression function satisfies conditions of
Thm.~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
m(r(\mathbf{x}; \theta_0, \theta_1))$,
for $m(r(\mathbf{x})) = ({1 + r(\mathbf{x})})^{-1}$, is monotonic with
$r(\mathbf{x}; \theta_0, \theta_1)$.
